---
title: Binning Numerical Features
author: Daniel S.
output:
  prettydoc::html_pretty:
    theme: tactile
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# References

- [Zheyuan Li, Simon N. Wood: "Faster model matrix crossproducts for large generalized linear models with discretized covariates"](https://link.springer.com/content/pdf/10.1007/s11222-019-09864-2.pdf)
- [Lang et al.: "Multilevel structured additive regression"](https://link.springer.com/article/10.1007/s11222-012-9366-0)
    - First formulation of _binning_ (page 8, 3.3. remak 2)


# Goal

"___Reducing memory load and runtime performance by shorten the number of observations used for fitting a linear model.___"


# Methodology

## Faster model matrix crossproducts with discretized covariates

The linear models requires the computation of the matrix $X^TWX$ where $X\in\mathbb{R}^{n\times p}$ is the design matrix and $W$ a weight matrix. We restrict our scenario here to just diagonal weight matrices and $X$ as base of a one feature. This does not necessarily mean that $X\in\mathbb{R}^{1\times n}$ but that $X$ is created as basis of a feature vector $x\in \mathbb{R}^n$ (for instance spline basis).

The core idea of Simon Wood and Zheyuan Li is to create another matrix $\bar{X}\in\mathbb{R}^{\bar{n}\times p}$ and $\bar{W}\in\mathbb{R}^{\bar{n}\times\bar{n}}$ based on the unique values of $x$ so that the following property holds:
$$
X^TWX = \bar{X}^T\bar{W}\bar{X}
$$

- index vector (key here)
- Algo to create W
- Also show how to compute $X^Ty$ in this system
- Small showcase in R

## Binning numerical features

The idea here is to discretize the vector $x\in\mathbb{R}^n$ to a vector $\bar{x}$ where the number of unique values $\text{unique}(\bar{x}) = m_u < n_u = \text{unique}(x)$. The vector $\bar{x}$ is then used for faster model matrix crossproducts as explained [above](#faster-model-matrix-crossproducts-with-discretized-covariates).

- How to best discretize?
- Small showcase in R

# Microbenchmark

```{r}
head(iris)
```

